Range Analysis of Whole Programs

### 2012-03-13 doc/Manuscript/paper.tex
Wrote the section about experiments.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.8\textwidth]{images/algorithm}
\end{center}
\caption{\label{fig:algorithm}
Our algorithm to solve the range analysis problem.
Rounded boxes are optional steps.}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\begin{eqnarray*}
\begin{array}{c}
\inferrule{\lb{I[Y]} = -\infty \\ \lb{e(Y)} > -\infty}
{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
\\
\\
\inferrule{\lb{I[Y]} > \lb{e(Y)}}
{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
\\
\\
\inferrule{\ub{I[Y]} = +\infty \\ \ub{e(Y)} < +\infty}
{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
\\
\\
\inferrule{\ub{I[Y]} < \ub{e(Y)}}
{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
\end{array}
\end{eqnarray*}
\end{center}
\caption{\label{fig:crop_analysis}Cousot and Cousot's narrowing operator.}
\end{figure}

@@

\noindent
\textbf{Structural Properties.}
Table~\ref{tab:struct} shows structural properties of the programs in
the SPEC 2006 benchmark suite.
Most of the strongly connected components that constitute
the constraint graphs are singletons.
For instance, 99.11\% of the SSCs found in gcc (\texttt{403.gcc}) have only
one node.
Therefore the most complex phases of our algorithm are
confined to a few parts of the constraint graphs that we have to handle.
Furthermore, the strongly connected components are not large -- our largest
component, found in \texttt{403.gcc} has 2,131 nodes, and exists due to a long
chain of mutually recursive function calls.
These two observations imply that our runtime is likely to be linear in
practice, as we show in Figure~\ref{fig:TimeCorr}.

\begin{table}[htdp]
\begin{center}
\begin{scriptsize}
\renewcommand{\tabcolsep}{0.1cm}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|} \hline
Name	    & 464 &	473 & 483 & 458 & 429 & 471 & 403 & 445 & 462 & 401 & 456 \\ \hline
\#I  & 139626 & 8900 & 593765 & 26947 & 2738 & 94036 & 887191 & 141330 &  6237 & 17663 & 61736 \\ \hline
\#$N_v$ & 97494 & 5490 & 352423 & 14407 & 1853	 & 50095 & 449442 & 84846 & 3293 & 12517 & 38409 \\ \hline
\#S  & 91731 & 5108 & 346451 & 13309 & 1751 & 48784 & 435372 & 79909 & 3003 & 11539 & 35619 \\ \hline
L    & 38 & 49 & 120 & 116 & 16 & 258 & 2131 & 311 & 9 & 18 & 41  \\ \hline
\#U  & 89461 & 4983 & 344132 & 12919 & 1726 & 48400 & 431703 & 78407 & 2885 & 11267 & 34631 \\ \hline
\end{tabular}
\end{scriptsize}
\end{center}
\caption{\label{tab:struct}Structural Properties of the programs in the SPEC CPU
2006 benchmark suite.
\#I: number of assembly instructions in the original program.
\#$N_v$: number of variable nodes in the constraint graph.
% TODO: add also number of constraint nodes.
\#S: number of strongly connected components (SCC).
L: size of the largest SCC.
\#U: number of SCCs with only one node.}
\end{table}%

@@

\noindent
\textbf{Runtime Behavior.}
Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
run our algorithm and the size of the input programs.
We show data for the 100 largest benchmarks in our test suite, in number
of variable nodes in the constraint graph.
We perform function inlining before running our analysis, to increase program
sizes.
Each point in the X line corresponds to a benchmark.
We analyze the smallest benchmark in this set, \texttt{Prolangs-C}, which has
1,131 variable nodes in the constraint graph, in 20ms.
We take 15,91secs to analyze our largest benchmark, \texttt{403.gcc}, which,
after function inlining, has 1,266,273 assembly instructions, and a
constraint graph with 679,652 variable nodes.
For this data set, the coefficient of determination $(R^2)$ is 0.967, which
provides very strong evidence about the linear asymptotic complexity of our
implementation.

@@

%Chart 1: shows complexity of the algorithm, plotting program size and time
% to run the analysis.

%Chart 2: time distribution for SPEC only

% Intra/Inter/Inline comparison
% =============================

% Chart 3: Time comparison between intra, inter and inlining:
% Each bar is inter/intra, or inline/intra
% We want to show that the whole program analysis is fast.

% (I) Chart 4: Precision comparison between intra/inter/inline
% *not for spec*: take five largest of bitwise, stanford and SPEC CINT.
% Add the three geomeans to the chart.

% (I) Chart 5: Same as (4), but this time precision in terms of [c, c], [-, c],
% [+, c], [-, +]. Perhaps take off [-, +]

>>>>

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.7\textwidth]{images/algorithm}
\end{center}
\caption{\label{fig:algorithm}
Our algorithm to solve the range analysis problem.
Rounded boxes are optional steps.}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\begin{eqnarray*}
\begin{array}{c@{\hspace{0.9cm}}c}
\inferrule{\lb{I[Y]} = -\infty \\ \lb{e(Y)} > -\infty}
{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
&
\inferrule{\lb{I[Y]} > \lb{e(Y)}}
{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
\\
\\
\inferrule{\ub{I[Y]} = +\infty \\ \ub{e(Y)} < +\infty}
{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
&
\inferrule{\ub{I[Y]} < \ub{e(Y)}}
{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
\end{array}
\end{eqnarray*}
\end{center}
\caption{\label{fig:crop_analysis}Cousot and Cousot's narrowing operator.}
\end{figure}

@@

\noindent
\textbf{Structural Properties.}
Table~\ref{tab:struct} shows structural properties of the programs in
the SPEC 2006 benchmark suite.
Most of the strongly connected components that constitute
the constraint graphs are singletons.
For instance, 99.11\% of the SSCs found in gcc (\texttt{403.gcc}) have only
one node.
Therefore the most complex phases of our algorithm are
confined to a few parts of the constraint graphs that we have to handle.
Furthermore, the strongly connected components are not large -- our largest
component, found in \texttt{403.gcc} has 2,131 nodes, and exists due to a long
chain of mutually recursive function calls.
We do not count control dependency edges when measuring the size of SCCs, as
they have no impact on the algorithm explained in Section~\ref{sub:micro}.
Finally, as we see in the table, any node was visited at most three times by
the narrowing operator.
These three observations imply that our runtime is likely to be linear in
practice, as we show in Figure~\ref{fig:TimeCorr}.

\begin{table}[htdp]
\begin{center}
\begin{scriptsize}
\renewcommand{\tabcolsep}{0.1cm}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|} \hline
Name	    & 464 &	473 & 483 & 458 & 429 & 471 & 403 & 445 & 462 & 401 & 456 \\ \hline
\#I  & 139626 & 8900 & 593765 & 26947 & 2738 & 94036 & 887191 & 141330 &  6237 & 17663 & 61736 \\ \hline
\#$N_v$ & 97494 & 5490 & 352423 & 14407 & 1853	 & 50095 & 449442 & 84846 & 3293 & 12517 & 38409 \\ \hline
\#$N_c$ & 69530 & 3799 & 213535 & 11164 & 1156 & 28943 & 309310 & 63506 & 2336 & 8731 & 24530 \\ \hline
\#S  & 91731 & 5108 & 346451 & 13309 & 1751 & 48784 & 435372 & 79909 & 3003 & 11539 & 35619 \\ \hline
L    & 38 & 49 & 120 & 116 & 16 & 258 & 2131 & 311 & 9 & 18 & 41  \\ \hline
\#U  & 89461 & 4983 & 344132 & 12919 & 1726 & 48400 & 431703 & 78407 & 2885 & 11267 & 34631 \\ \hline
\#V  & 3	2 & 2 & 2 & 2 & 2 & 2 & 3 & 2 & 2 & 2 \\ \hline
\end{tabular}
\end{scriptsize}
\end{center}
\caption{\label{tab:struct}Structural Properties of the programs in the SPEC CPU
2006 benchmark suite.
\#I: number of assembly instructions in the original program.
\#$N_v$: number of variable nodes in the constraint graph.
\#$N_c$: number of constraint nodes in the constraint graph.
% TODO: add also number of constraint nodes.
\#S: number of strongly connected components (SCC).
L: size of the largest SCC.
\#U: number of SCCs with only one node.
\#V: maximum number of times any variable node was visited by narrowing.}
\end{table}%

@@

\noindent
\textbf{Asymptotic complexity.}
Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
run our algorithm and the size of the input programs.
We show data for the 100 largest benchmarks in our test suite, in number
of variable nodes in the constraint graph.
We perform function inlining before running our analysis, to increase program
sizes.
Each point in the X line corresponds to a benchmark.
We analyze the smallest benchmark in this set, \texttt{Prolangs-C}, which has
1,131 variable nodes in the constraint graph, in 20ms.
We take 15,91secs to analyze our largest benchmark, \texttt{403.gcc}, which,
after function inlining, has 1,266,273 assembly instructions, and a
constraint graph with 679,652 variable nodes.
For this data set, the coefficient of determination $(R^2)$ is 0.967, which
provides very strong evidence about the linear asymptotic complexity of our
implementation.

@@

\noindent
\textbf{Time distribution.}
Figure~\ref{fig:timeComp}(Left) shows how the processing time is distributed
among the main phases of the inter-procedural version of our algorithm.
In this case, we have inlined functions to increase program sizes.
We analyze the entire SPEC 2006 integer benchmarks in 41.42secs.
The time to build the e-SSA representation ss 7.03secs; the time to extract
constraints from the program and build the constraint graphs is 7.87secs;
the time to run Nuutila's algorithm to separate the constraint graph into
strong components is 10.55secs, and the time to run the analysis itself is
15.96secs.
Notice that the Nuutila's entry includes also the time to sort the SCCs
topologically.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.48\textwidth]{images/timeDistr}
\includegraphics[width=0.48\textwidth]{images/timeIntraInterInline}
\end{center}
\caption{\label{fig:timeComp}
(Left) Time distribution across different phases of our inter+inline
algorithm.
(Right) Runtime comparison between the intra, inter e inter+inline versions of
our algorithm.
The bars are normalized to the time to run the intra-procedural version.
}
\end{figure}

\noindent
\textbf{Intra vs Inter-procedural runtimes.}
As we saw in Figure~\ref{fig:algorithm}, our implementation supports three
different execution modes: intra-procedural, inter-procedural, and
inter-procedural with inlining.
The latter approach gives us some form of context-sensitiveness, because our
baseline compiler, LLVM, is able to inline non-recursive function calls.
Figure~\ref{fig:timeComp}(Right) compares these three different setups.
All the times are normalized to the time to run the intra-procedural version.
On the average, the intra-procedural mode is XX\% faster than the
inter-procedural mode.
If we perform function inlining, then this difference is larger: XX\%.

\noindent
\textbf{The runtime impact of the e-SSA intermediate representation.}
Figure~\ref{fig:eSSA_data} shows how the e-SSA transformation impacts the
runtime of our algorithm.
As we see in Figure~\ref{fig:eSSA_data}(Left), this alternative representation
increases the size of the SPEC 2006 integer benchmarks by a factor of XX\%,
when compared to the original program, in SSA form.
Such growth is due to the $\sigma$-functions used to split live ranges
after conditionals, and to the extra $\phi$-functions necessary to preserve the
single static assignment property.
The e-SSA form increases the runtime of our analysis by XX\%, as we see in
Figure~\ref{fig:eSSA_data}(Right).
In that figure we are comparing only the time to perform range analysis.
The bars do not include the time to perform the e-SSA conversion, which we
show in Figure~\ref{fig:timeComp}(Left).

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.48\textwidth]{images/size_e_SSA}
\includegraphics[width=0.48\textwidth]{images/time_e_SSA}
\end{center}
\caption{\label{fig:eSSA_data}
(Left) Program growth due to the conversion from the SSA to the e-SSA form.
(Right) Runtime slowdown of the range analysis when running it on the e-SSA
form programs relative to the time to run it on the SSA-form programs.
}
\end{figure}

\noindent
\textbf{The impact of whole program analysis on precision.}
Figure~\ref{fig:precision} shows that the whole program analysis increases
moderately the precision of our analysis.
We show results for the five largest programs in three different categories of
benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our initial goal when developing this analysis was to support a bitwidth-aware
register allocator.
Thus, we measure precision in terms of the average number of bits that our
analysis allows us to save to store the program variables.
Our results for the SPEC programs were disappointingly: on the average, the
intra-procedural version of our analysis saves 5.23\% of the total program
bitwidth.
By using the inter-procedural version, with function inlining to simulate
context-sensitiveness, we can increase this number to 8.89\%.
A manual inspection of the SPEC programs reveal that this result is expected:
such benchmarks manipulate files, and their code do no provide enough
explicit constants to power our analysis.
However, with numerical benchmarks we fare much better.
On the average our inter procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
Notice that this number is better than the one we get with inlining (31.97\%),
because functions in those benchmarks are called only at one location.
Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is brought down by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which cannot be reduced.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
Our results are on par with those found by the original authors.
Notice that all these benchmarks contain only the \texttt{main} function; thus,
there is no different between the different versions of our algorithm.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/precision}
\end{center}
\caption{\label{fig:precision}
The impact of whole program analysis on precision.
Each bar measures the average bitwidth reduction after we run our analysis.}
\end{figure}

\noindent
\textbf{The impact of e-SSA on precision.}
Figure~\ref{fig:essaDivSSA} shows that the e-SSA transformation improves the
precision of our analysis dramatically.
If we convert the input programs from the original SSA representation used
by LLVM to the e-SSA format, then, in some cases, such as in
\texttt{Stanford.queens}, in the intra-procedural case with function
inlining, we can increase the bitwidth reduction by 154x.
This impressive difference happens because the standard SSA representation does
not give our analysis any subsidy to acquire information from the outcome of
conditionals.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/essaDivSSA}
\end{center}
\caption{\label{fig:essaDivSSA}
The impact of the e-SSA transformation on precision.
Each bar gives the ratio (bitwidth reduction on e-SSA form)/(bitwidth reduction
on SSA form).}
\end{figure}
